---
title: "EDA: Mortgage Lending Patterns Across U.S. Regions"
format: pdf
editor: visual
---

## Introduction

This exploratory analysis investigates potential regional and demographic patterns in U.S. mortgage approvals using a sample of the 2023 HMDA dataset. Our aim is to identify whether there are observable disparities—possibly indicating digital redlining—across the **West Coast (CA)**, **East Coast (NY)**, and the **South (GA)**. I focused on loan-level characteristics and demographic features and will prepare for a model-based clustering analysis.

------------------------------------------------------------------------

## Data Setup

```{r}
#| message: false
library(tidyverse)

# Load the data
hmda <- read_csv("HMDA_clean.csv")

# Filter relevant states
target_states <- c("CA", "NY", "GA")

# Subset to relevant columns and filter for the three regions
hmda_sub <- hmda %>%
  filter(state_code %in% target_states) %>%
  select(state_code, loan_amount, income, approved,
         derived_race, derived_sex, derived_ethnicity,
         tract_minority_population_percent, tract_to_msa_income_percentage) %>%
  drop_na()

# Create a region variable
hmda_sub <- hmda_sub %>%
  mutate(region = case_when(
    state_code == "CA" ~ "West",
    state_code == "NY" ~ "East",
    state_code == "GA" ~ "South"
  ))
```

------------------------------------------------------------------------

## Univariate EDA

## Additional Variable Distributions

### Race Distribution

```{r}
ggplot(hmda_sub, aes(x = derived_race)) +
  geom_bar(fill = "steelblue") +
  theme_minimal() +
  labs(title = "Distribution of Race", x = "Race", y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

*The dataset is dominated by White applicants, with notable representation of Asian and Black or African American applicants. A substantial number of applications lack reported race information ("Race Not Available"). Smaller minority groups, such as American Indian/Alaska Native and Native Hawaiian/Pacific Islander, are underrepresented in the data.*

### Sex Distribution

```{r}
ggplot(hmda_sub, aes(x = derived_sex)) +
  geom_bar(fill = "darkgreen") +
  theme_minimal() +
  labs(title = "Distribution of Sex", x = "Sex", y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

*The dataset contains roughly equal proportions of male and joint applicants, with slightly fewer female applicants. A substantial number of records also fall under "Sex Not Available," indicating some missing or unreported gender data.*

### Ethnicity Distribution

```{r}
ggplot(hmda_sub, aes(x = derived_ethnicity)) +
  geom_bar(fill = "purple") +
  theme_minimal() +
  labs(title = "Distribution of Ethnicity", x = "Ethnicity", y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

*Most applicants identify as "Not Hispanic or Latino." There is a meaningful presence of "Hispanic or Latino" applicants, as well as a large number of records where ethnicity is "Not Available," which may affect analysis of potential ethnicity-based disparities.*

### Region Distribution (Counts)

```{r}
ggplot(hmda_sub, aes(x = region)) +
  geom_bar(fill = "orange") +
  theme_minimal() +
  labs(title = "Distribution of Region", x = "Region", y = "Count")
```

*The dataset includes a disproportionately larger number of applications from the West (California), compared to the East (New York) and South (Georgia). This imbalance reflects both differences in population size and potentially differing volumes of mortgage activity across regions.*

***Because both income and loan amounts are highly skewed, I need to visualize them on a log scale to better illustrate differences across applicants.***

```{r}
# Loan Amount
ggplot(hmda_sub, aes(x = loan_amount)) +
  geom_histogram(bins = 50, fill = "#1f77b4") +
  scale_x_log10() +
  labs(title = "Distribution of Loan Amounts (Log Scale)", 
       x = "Loan Amount (log10)", y = "Count")

# Income
ggplot(hmda_sub, aes(x = income)) +
  geom_histogram(bins = 50, fill = "#2ca02c") +
  scale_x_log10() +
  labs(
    title = "Distribution of Income (Log Scale)", 
    x = "Income (log10)", y = "Count")

# Approval Rate
hmda_sub %>%
  count(approved) %>%
  mutate(rate = n / sum(n)) %>%
  ggplot(aes(x = factor(approved), y = rate)) +
  geom_col(fill = "#ff7f0e") +
  labs(title = "Approval Rate", x = "Approved (1 = Yes)", y = "Proportion")
```

------------------------------------------------------------------------

## Bivariate EDA

```{r}
# Loan Amount by Race
ggplot(hmda_sub, aes(x = derived_race, y = loan_amount)) +
  geom_boxplot() +
  labs(title = "Loan Amount by Race", x = "Race", y = "Loan Amount")

# Approval Rate by Region
hmda_sub %>%
  group_by(region) %>%
  summarise(approval_rate = mean(approved)) %>%
  ggplot(aes(x = region, y = approval_rate, fill = region)) +
  geom_col() +
  labs(title = "Approval Rate by Region", x = "Region", y = "Approval Rate")
```

------------------------------------------------------------------------

## Multivariate Relationships

```{r}
# Loan Amount vs Income colored by approval
ggplot(hmda_sub, aes(x = income, y = loan_amount, color = factor(approved))) +
  geom_point(alpha = 0.5) +
  labs(title = "Loan Amount vs Income by Approval", color = "Approved")

# Minority population percent by approval
ggplot(hmda_sub, aes(x = tract_minority_population_percent, fill = factor(approved))) +
  geom_histogram(position = "identity", alpha = 0.6, bins = 40) +
  labs(title = "Tract Minority % by Approval Status", x = "Minority Population %", fill = "Approved")
```
