---
title: "EDA: Mortgage Lending Patterns Across U.S. Regions"
format: html
editor: visual
---

## Introduction

This exploratory analysis investigates potential regional and demographic patterns in U.S. mortgage approvals using a sample of the 2023 HMDA dataset. Our aim is to identify whether there are observable disparities—possibly indicating digital redlining—across the **West Coast (CA)**, **East Coast (NY)**, and the **South (GA)**. I focused on loan-level characteristics and demographic features and will prepare for a model-based clustering analysis.

------------------------------------------------------------------------

## Data Setup

```{r}
#| message: false
library(tidyverse)

# Load the data
hmda <- read_csv("HMDA_clean.csv")

# Filter relevant states
target_states <- c("CA", "NY", "GA")

# Subset to relevant columns and filter for the three regions
hmda_sub <- hmda %>%
  filter(state_code %in% target_states) %>%
  select(state_code, loan_amount, income, approved,
         derived_race, derived_sex, derived_ethnicity,
         tract_minority_population_percent, tract_to_msa_income_percentage) %>%
  drop_na()

# Create a region variable
hmda_sub <- hmda_sub %>%
  mutate(region = case_when(
    state_code == "CA" ~ "West",
    state_code == "NY" ~ "East",
    state_code == "GA" ~ "South"
  ))
```

------------------------------------------------------------------------

## Univariate EDA

***Because both income and loan amounts are highly skewed, I need to visualize them on a log scale to better illustrate differences across applicants.***

```{r}
# Loan Amount
ggplot(hmda_sub, aes(x = loan_amount)) +
  geom_histogram(bins = 50, fill = "#1f77b4") +
  scale_x_log10() +
  labs(title = "Distribution of Loan Amounts (Log Scale)", 
       x = "Loan Amount (log10)", y = "Count")

# Income
ggplot(hmda_sub, aes(x = income)) +
  geom_histogram(bins = 50, fill = "#2ca02c") +
  scale_x_log10() +
  labs(
    title = "Distribution of Income (Log Scale)", 
    x = "Income (log10)", y = "Count")

# Approval Rate
hmda_sub %>%
  count(approved) %>%
  mutate(rate = n / sum(n)) %>%
  ggplot(aes(x = factor(approved), y = rate)) +
  geom_col(fill = "#ff7f0e") +
  labs(title = "Approval Rate", x = "Approved (1 = Yes)", y = "Proportion")
```

------------------------------------------------------------------------

## Bivariate EDA

```{r}
# Loan Amount by Race
ggplot(hmda_sub, aes(x = derived_race, y = loan_amount)) +
  geom_boxplot() +
  labs(title = "Loan Amount by Race", x = "Race", y = "Loan Amount")

# Approval Rate by Region
hmda_sub %>%
  group_by(region) %>%
  summarise(approval_rate = mean(approved)) %>%
  ggplot(aes(x = region, y = approval_rate, fill = region)) +
  geom_col() +
  labs(title = "Approval Rate by Region", x = "Region", y = "Approval Rate")
```

------------------------------------------------------------------------

## Multivariate Relationships

```{r}
# Loan Amount vs Income colored by approval
ggplot(hmda_sub, aes(x = income, y = loan_amount, color = factor(approved))) +
  geom_point(alpha = 0.5) +
  labs(title = "Loan Amount vs Income by Approval", color = "Approved")

# Minority population percent by approval
ggplot(hmda_sub, aes(x = tract_minority_population_percent, fill = factor(approved))) +
  geom_histogram(position = "identity", alpha = 0.6, bins = 40) +
  labs(title = "Tract Minority % by Approval Status", x = "Minority Population %", fill = "Approved")
```

------------------------------------------------------------------------

## Clustering Motivation

From the visual analysis above, I see suggestive patterns involving **income**, **loan amount**, **approval**, and **demographics** (especially race and minority neighborhood composition). For my midterm, I will apply **model-based clustering** using variables such as:

-   `loan_amount`, `income`
-   `tract_minority_population_percent`
-   `approval` (for validation only)

to investigate whether **unobserved clusters** align with protected class variables—possibly indicating algorithmic bias or digital redlining.
